Agile Formality: A \mole of\\Software Engineering Practices

### 2010-08-24 papers/AgileFormality/paper.tex
Revised sections 1, 5, and 6.

\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STs}{\emph{SenseTiles}\xspace}
\newcommand{\SB}{\emph{Sensor Board}\xspace} 
\newcommand{\STSB}{\ST \SB\xspace} 
\newcommand{\STSBs}{\ST \emph{Sensor Boards}\xspace}
\newcommand{\STPU}{\ST \emph{Processor Unit}\xspace}
\newcommand{\STPUs}{\ST \emph{Processor Units}\xspace}
\newcommand{\STU}{\ST \emph{Unit}\xspace} 
\newcommand{\STUs}{\ST \emph{Units}\xspace}

@@

\title{Agile Formality: A ``Mole'' of Software Engineering Practices}

\author{Vieri del Bianco and Dragan Stosic\\
  University College Dublin, Ireland\\
  \texttt{vieri.delbianco@ucd.ie} and \texttt{dragan.stosic@gmail.com}\\
  Joseph R.~Kiniry \\
  IT University of Copenhagen, Denmark\\
  \texttt{kiniry@itu.dk}
}

@@

Members of the agile programming and formal methods communities do not always see face-to-face.  
These two communities often do not talk to or learn from each other.  
Only recently, as highlighted by the September 2009 issue of IEEE Software, the IFIP workshop on balancing agility and formalism in software engineering, and the first edition of the international workshop for formal methods and agile methods, ideas from the two communities begun to be synthesized.  
While the problem-solving approaches and psychological attitudes of members of the two communities differ widely, we exploit this clash of viewpoints, exploiting a new development processes that actually blends, rather than mashes together, best practices from the two worlds.  
This paper summarizes our process and a supporting complex case study, showing that it is not only possible, but tasty, to combine the ``chili pepper'' of formal methods and the ``chocolate'' of agile programming, thus producing a tasty ``Mole'' (as in the highly-spiced Mexican sauce) of software engineering practices.

@@

% FM
Formal methodologists often favour an in-depth, think-first approach, where the problem is understood, formalized, and solved; usually, but not always, adopting a waterfall-style of development.  
Once a formalization is developed, development and verification proceed, in an automated and interactive fashion.  
Consequently, projects that use formal methodologies (FMs) typically focus on critical systems with fixed requirements and somewhat flexible deadlines (``We'll ship it when it is right.'').

% AM
Agile methodologists favour an highly incremental and iterative approach, specifically tailored to cope with changing requirements and precise deadlines.  
In projects that use agile methodologies (AMs), it is often the case that the problem is only partially understood, focusing only on the aspects that must be implemented in the current development iteration.  
The solution initially developed is usually not optimal, but as it is refined through continuous code refactorings, its quality and validity improve.  
Test suites are used for several purposes: the most important are system verification, system documentation, and requirement specification, and they are strictly handmade.  
AMs are typically used in development settings where changing requirements and rapid delivery of the product are paramount.

@@

% tests terminology
Tests are usually classified by dichotomies: unit vs. functional, black-box vs. white-box, testing vs. design, tester vs. coder. 
Yet testing doesn't always fit in the dichotomies, especially in the case of agile testing approaches~\cite{Beck2007}.
Black-box vs. white-box dichotomy is particularly problematic, since the box boundaries change the perspective.
In the paper, functional tests are always black-box tests, coverage tests are always white-box tests, generated tests are always black-box tests (the box is the module specified); other strict rules do not hold.

% development terminology
Additionally, ``...development of a system,'' more precisely means ``...analysis, design, development, verification, formal validation of a system''.

@@

Proposing new practices with no supporting evidence is an uninteresting proposition.  
A complex case study is mandatory to test bench the proposed practices.  
Our case study focuses on the development of a device driver to control and communicate with an embedded custom circuit board equipped with more than a dozen sensors.
The communication protocol with the board is asynchronous and packet-based and the board is novel and custom-designed.  
The protocol is focused on controlling the board and reporting sensor measurements and board status.

This software driver has been developed using our process while statisfying seemingly contradictory requirements: (1)~the protocol implementation must be thoroughly verified; (2)~parts of the protocol must be formally specified; (3)~a board simulator must be designed, implemented, and verified to match the specifications of, and actual behaviour of, the physical circuit board; (4)~if board components change during development, the associated formal specifications and the simulator whose behavior reflects such changes must be updated accordingly; (5)~strict software delivery deadlines.  
Consequently, these process requirements constitute an environment that benefits from both agile and formal approaches to software development.

A novel development process has been drafted to cope with the contrasting requirements of the case study, blending agile and formal practices.  
In the following the development process and the practices are summarized, and their successful application to the case study is reported, yielding several promising results.

@@

% what is needed, high-level
To enable the integration of the two worlds, the highly iterative and incremental approach found in most of the AMs must be maintained, and formal validation methodologies, traditionally used in a waterfall development process (formally specify the system, implement the system, validate the implementation against the specification), must be adapted to a highly iterative one.

% what must be considered: TDD
The most common software verification practices in AMs must be considered, when trying to integrate FMs and AMs, since they are a fundamental part of the development process. 
The most significant verification related practice is Test Driven Development (TDD)~\cite{Beck2003}: TDD is a software development technique, originally defined in Extreme Programming (XP)~\cite{Beck2004} methodology, characterized by writing unit tests before writing the code.

% test driven kinds
The test driven approach relies on writing the tests before the actual implementation. 
It is applicable to a very small scale (TDD) yielding to activity cycles as short as a few minutes.
It is applicable to a large scale as well (requirements immediately translated into functional tests), yielding to activity cycles as long as a full delivery iteration (usually no longer than a couple of months).
The test driven approach is a cornerstone of XP, but it is so popular that has been adopted in many other AMs as well.
It is also considered a good software development technique when used on its own, regardless of the enclosing development process.

% previous related work
Some tentative attempts to reconcile FMs and AMs have been developed.
The recurrent approach of the attempts consists in removing the handmade test suites, substituting them with different kinds of automated formal validation based on the system formal specification. 
The automated validation can be as simple as enabling runtime verification of assertions or automatically generating a test suite based on the system properties, or as complex as actually proving the whole system behaviour through static model and code checkers or theorem provers (formal verification).

% a quick analysis of the related works
Eleftherakis and Cowling in~\cite{Eleftherakis2003} propose XFun, an iterative and incremental process to generate X-Machines models, an extension of Finite State Machines. 
The development process impose a complete specification of the system, and its verification is done exclusively using the tools provided by the X-Machines FM. 
The development process proposed in~\cite{Herranz2003b} by Herranz and Moreno-Navarro is quite similar: they propose an iterative process to model a system using SLAM-SL, an object-oriented specification language, and its tools. 
Some XP practices are fully adopted or considered compatible, as pair-programming, iterative and incremental development, system refactoring; while tests are completely replaced by the verification tools provided by the SLAM suite.
A different FM, but same approach, is found also in~\cite{Suhaib2005} where XFM methodology is explained.

% previous work: a common approach
As already mentioned, in all these attempts, the test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at minimum, a specification of all the relevant aspects of the system under development, to enable simple verification techniques such as runtime verification (assertions and design by contract~\cite{Meyer1997} preconditions, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
Whether a complete and sound system specification, supported by an appropriate formal language (to specify model and properties) and a constrained programming language, can be able to statically prove properties about the system~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

% previous work: why is not satisfying
The approach is simple: the verification, documentation and requirement specification purposes of the tests are moved on to the formal specifications, and a complete formal specification of the system under development is needed to apply these methodologies.
The problem is that this requirement is hard to fulfil, since the effort required to write a complete formal specification of the system, in most real world complex cases, is usually greater than writing a suite of tests~\cite{Glass2002}.
If a complete formal specification is not built, all the previous methodologies share a common problem: the parts informally specified are not verified at all.
These can obviously be verified with the traditional methods, but the loops between development artifacts and related activities (code, tests, design, refactoring) are neither detailed nor enforced.
The loops and relations between the different development artifacts are the inner engine of many AMs, they impose an iterative and incremental pace to the development process.

% outsider
Liu in~\cite{Liu2010} takes a different approach where SOFL methodology~\cite{Liu2004} is merged into agile processes. 
SOFL methodology is based on a three step specification approach: (1) informal specification, (2) semi-formal specification, (3) formal specification. 
The agile adaptation consists mainly in shorter loops between activities, and into a reduced set of formal specifications.
The formal specifications are used only to help understanding ambiguous statements in the semi-formal specification, and are not maintened while the system evolves; testing and inspections are based on the semi-formal specification.
The formal specification are partial, which is interesting, but used only for documentation purposes and than discarded, they are not used at all in the verification of the system.

% what we want to achieve
The objectives we want to achieve are the following ones: to formally specify only parts of the system, in order to cope with constrained resources and unstable requirements; to draft an highly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in AMs (especially in XP), to solve this problem in an highly iterative development process.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how handmade and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal validation and agile verification practices: a ``Mole'' of verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly spiced Mexican sauce made chiefly from chili peppers (agile) and chocolate (formal), served with meat (working system).}.

@@

``Mole'' practices are applied in the context of a real world case study that matches the problem requirements previously detailed: unstable requirements and development artifacts that need to be formally specified and validated.

@@

% device drivers and formal specification
When considering the development of the device and of its software driver as separate entities that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware-software co-design~\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}, all of which focus on a specification that aims to be as complete as possible.
A complete specification is neither feasible nor convenient in our case. 
The protocol specification must be enhanced incrementally and its complete specification cannot be provided.  
This makes the problem an ideal candidate to test our ``Mole'' of practices.

@@

\subsection{``Mole'' practices}

@@

% all together blends into a mole
When the ``Mole'' verification practices are used together the initial objectives are achieved: an incremental and iterative process, where development pace is defined by very short development cycles similar to TDD; the complete freedom to decide what is formally specified; the development cycles and the associated verification practices maintain the corresponding artifacts involved consistent, the correct development cycle is determined by the formality of the artifacts involved, ranging from pure development cycles, shown in Figure \ref{fig:tdd_diagram} and \ref{fig:fdd_diagram}, to blended ones, shown in Figure \ref{fig:fdd_hand_made_tests_diagram} and \ref{fig:tdf_diagram}.

% =====================================================================
\section{``Mole'' verification practices applied}
\label{sec:a_complex_case_study}

\subsection{Data stream, protocol description and protocol specification}
\label{subsec:data_stream_and_protocol_description}

@@

% verification
Verification is provided by handmade test suite, JML annotations and generated unit test suite. 
It has been built starting from handmade tests only (TDD development cycles).
Later on JML annotations have been added incrementally, using a mix of ``Mole'' practices: to move from handmade tests to formal specifications, TDF have been initially used; than all the ``Mole'' practices have been used according to the artifacts involved.

% development cycles
The test suites are used to verify board simulators (high-level), RAC and JML annotations are used to check the formally specified properties at runtime.
Code and tests have been initially implemented using TDD, than using all the ``Mole'' verification practices.

@@

\subsection{JML specification examples}
\label{subsec:a_jml_specification_example}

@@

\subsection{Test cases}
\label{subsec:test_cases}

@@

\subsection{Test cases on the job}
\label{subsec:test_cases_on_the_job}

@@

\section{Retrospective on ''Mole`` practices effectiveness}
\label{subsec:test_cases_retrospectives}

% preconditions on projects
''Mole`` practices are very easy to be applied and work with.
Many projects could benefit from the practices, not only the unstable ones that requires a certain level of formality.
Even projects with no formal specification requirements at all could benefit from the practices, since some system properties are easy to specify and could actually reduce the test suites development effort.

% precondition on formal methodology and associated formal suite
The cornerstone of the practices is the FM with its supporting tools, and the precondition on the tools are not easy to be fulfilled.
Methodology that require a complete formalization are impossibile to be applied.
Beside, it is not clear if a tool suite that does not provide a run time verification environment could be used succesfully, not it is clear whether formal methods not based on the contract paradigma (preconditions and postconditions) could be applied at all.
These are open questions that will nedd to be addressed in future works to assess the real appliability of the ''Mole`` practices.

% when to use what
There no strict rules to determine which of the ''Mole`` practices to use to face a specific problem, but only advices.
TDD if we don't need anu specification, FDD with generated tests if the specification are not too difficult and the scenario is simple enough to be covered by generated tests, FDD with hand made tests if the scenarios is complex, TDF when the specification is complex.
The experience gathered is yet not enogh to distill idioms or patterns.

% what is working and what is not
FDD with hand made tests is amazing, it is as natural to work with as TDD. 
FDD with generated tests can be annoying because of the time needed to run the test suite, often only part of the test suite is executed so that the development loops take not too long.
TDF is great with complex preconditions, but really difficult with postconditions, especially if the soecification is made on top of already existing code. 
When developing postconditions after code, the test is not expected to fail, hence the feedback is limited and the specification could not be as strict as it was needed. 
In this specific case TDF effectiveness is limited.

@@

% the problem
In this paper we focus on a specific set of development problems that neither agile methodologies nor FMs are completely confortable to cope with.
The problems we are interested in are characterized by unstable requirements and artifacts which must be formally specified and verified, and are constrained by deadlines that cannot be postponed.
We propose a blend of agile and formal engineering practices, enclosed by an iterative and incremental development process: the ``Mole'' development process.

% Mole
The verification practices proposed recreate the fast feedback development environment we find in Test Driven Development, used with formal and informal artifacts.
A total of four practices are presented, two of them are conservative, following closely the development practices of FMs and AMs, while the other pair are innovative, blending together elements from both worlds.

% Mole at work
The practices have been applied succesfully to develop a driver for a custom embedded sensor board, equipped with board simulators and protocol verifiers. 
The protocol and the board specifications were informal and unstable. 
Yet a working software product was needed as soon as possible, with incrementally added functionality, and keeping it linked with the last specifications of the protocol and the board.
``Mole'' verification practices enabled us to keep in sync formal specifications (JML annotations), informal specifications (test suites) and source code, through very fast and short development cycles.
We developed handmade test suites, generated test suites and JML annotations, wich succesfully supported the verification of the driver and the simulators, and the board as well, when it was delivered.

% extending Mole
The practices are very general, and are applicable to different development languages and FMs.
One important constraint is on the FM, that must be able to cope with partial specifications; another constraint is on the tools available, that must be able to verify the system properties at runtime and partially automate the verification process of the formalized properties.

@@

>>>>

\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STs}{\emph{SenseTiles}\xspace}
\newcommand{\SB}{\emph{Sensor Board}\xspace} 
\newcommand{\STSB}{\ST \SB\xspace} 
\newcommand{\STSBs}{\ST \emph{Sensor Boards}\xspace}
\newcommand{\STPU}{\ST \emph{Processor Unit}\xspace}
\newcommand{\STPUs}{\ST \emph{Processor Units}\xspace}
\newcommand{\STU}{\ST \emph{Unit}\xspace} 
\newcommand{\STUs}{\ST \emph{Units}\xspace}
\newcommand{\mole}{``\emph{Mole}''\xspace}

@@

\title{Agile Formality: A \mole of\\Software Engineering Practices}

\author{Vieri del Bianco and Dragan Stosic\\
  University College Dublin, Ireland\\
  \texttt{vieri.delbianco@ucd.ie} and \texttt{dragan.stosic@gmail.com}\\
  \\
  Joseph R.~Kiniry \\
  IT University of Copenhagen, Denmark\\
  \texttt{kiniry@itu.dk}
}

@@

  Members of the agile programming and formal methods communities do
  not always see eye-to-eye.  These two communities often do not talk
  to or learn from each other.  Only recently, as highlighted by the
  September 2009 issue of IEEE Software, the IFIP workshop on
  balancing agility and formalism in software engineering, and the
  first edition of the international workshop for formal methods and
  agile methods, ideas from the two communities begun synthesize.
  While the problem-solving approaches and psychological attitudes of
  members of the two communities differ widely, we exploit this clash
  of viewpoints, creating a new development processes that actually
  blends, rather than mashes together, best practices from the two
  worlds.  This paper summarizes our process and a supporting complex
  case study, showing that it is not only possible, but tasty, to
  combine the ``chili pepper'' of formal methods and the ``chocolate''
  of agile programming, thus producing a tasty \mole (as in the
  highly-spiced Mexican sauce) of software engineering practices.

@@

% FM
Formal methodologists often favour an in-depth, think-first approach,
where the problem is understood, formalized, and solved; usually, but
not always, adopting a waterfall-style of development.  Once a
formalization is developed, development and verification proceed in an
automated and interactive fashion.  Consequently, projects that use
formal methodologies (FMs) typically focus on critical systems with
fixed requirements and somewhat flexible deadlines (``We'll ship it
when it is right.'').

% AM
Agile methodologists favour an highly incremental and iterative
approach, specifically tailored to cope with changing requirements and
precise deadlines.  In projects that use agile methodologies (AMs), it
is often the case that the problem is only partially understood, and
there is a focus only on the aspects that must be implemented in the
current development iteration.  The solution initially developed is
usually not optimal, but as it is refined through continuous code
refactorings, its quality and validity improve.  Test suites are used
for several purposes, the most important of which are system
verification, system documentation, and requirement specification, and
they are strictly handmade.  AMs are typically used in development
settings where changing requirements and rapid delivery of the product
are paramount.

@@

% tests terminology
Tests are usually classified by dichotomies: unit vs. functional,
black-box vs. white-box, testing vs. design, tester vs. coder.  Yet
testing does not always fit into these dichotomies, especially in the
case of agile testing approaches~\cite{Beck2007}.  The black-box
vs. white-box dichotomy is particularly problematic, since box
boundaries change one's perspective.  In this paper, functional tests
are always black-box tests, coverage tests are always white-box tests,
generated tests are always black-box tests (the box is the module
specified); other strict rules do not hold.

% development terminology
Additionally, ``...development of a system,'' more precisely means
``...analysis, design, development, verification, formal validation of
a system.''

@@

Proposing new practices with no supporting evidence is an
uninteresting proposition.  A complex case study is mandatory to
test-bench the proposed practices.  Our case study focuses on the
development of a device driver to control and communicate with an
embedded custom circuit board equipped with more than a dozen sensors.
The communication protocol with the board is asynchronous and
packet-based and the board is novel and custom-designed.  The protocol
is focused on controlling the board and reporting sensor measurements
and board status.

This software driver has been developed using our process while
satisfying seemingly contradictory requirements: (1)~the protocol
implementation must be thoroughly verified; (2)~parts of the protocol
must be formally specified; (3)~a board simulator must be designed,
implemented, and verified to match the specifications of, and actual
behaviour of, the physical circuit board; (4)~if board components
change during development, the associated formal specifications and
the simulator whose behavior reflects such changes must be updated
accordingly; (5)~there are strict software delivery deadlines.
Consequently, these process requirements constitute an environment
that benefits from both agile and formal approaches to software
development.

A novel development process has been drafted to cope with the
contrasting requirements of the case study, blending agile and formal
practices.  In the following, the development process and the practices
are summarized, and their successful application to the case study is
reported, yielding several promising results.

@@

% what is needed, high-level
To enable the integration of the two worlds, the highly iterative and
incremental approach found in most of the AMs must be maintained, and
formal validation methodologies, traditionally used in a waterfall
development process (formally specify the system, implement the
system, validate the implementation against the specification) must
be adapted to a highly iterative one.

% what must be considered: TDD
The most common software verification practices in AMs must be
considered when trying to integrate FMs and AMs since they are a
fundamental part of the development process.  The most significant
verification related practice is Test Driven Development
(TDD)~\cite{Beck2003}.  TDD is a software development technique,
originally defined in the Extreme Programming (XP)~\cite{Beck2004}
methodology and is characterized by a process that focuses on writing
unit tests for a module before writing the code of said module.

% test driven kinds
The test driven approach relies on writing tests before
implementations.  It is applicable to a scall and large-scale cycles.
In the former case, very small scale TDD yields activity cycles as
short as a few minutes.  In the latter, requirements are immediately
translated into functional tests resulting in activity cycles as long
as a full delivery iteration, which is usually no longer than a couple
of months.  The test driven approach is a cornerstone of XP, but it is
so popular that has been adopted in many other AMs as well.  It is
also considered a good software development technique when used on its
own, as it pre-dates XP and AMs considerably, regardless of the
enclosing development process.

% previous related work
Some tentative attempts to reconcile FMs and AMs have been developed
by others.  The reoccuring theme of these attempts is that idea that
handmade test suites are no longer used, but instead are replaced by
different kinds of automated formal validation based on a system-level
formal specification.  Automated validation can be as simple as
enabling runtime checking of assertions or automatically generating a
test suite based on the system properties, or as complex as proving
the whole system's behaviour through formal verification.

% a quick analysis of the related works
Eleftherakis and Cowling in~\cite{Eleftherakis2003} propose XFun, an
iterative and incremental process to generate X-Machines models, an
extension of Finite State Machines.  The development process impose a
complete specification of the system, and its verification is done
exclusively using the tools provided by the X-Machines FM.  The
development process proposed by Herranz and
Moreno-Navarro~\cite{Herranz2003b} is quite similar.  They propose an
iterative process to model a system using SLAM-SL, an object-oriented
specification language supported by tools.  Some XP practices are
fully adopted or considered compatible, like pair-programming,
iterative and incremental development, and system refactoring; while
tests are completely replaced by the verification tools provided by
the SLAM suite.  A different FM, but similar approach, is found also
in the work of Suhaib~\cite{Suhaib2005} where the XFM methodology is
used.

% previous work: a common approach
As already mentioned, in all of these attempts, test suites are
automatically generated using the system formal specification, or are
completely replaced by formal verification, typically through the use
of theorem provers and/or static checkers.  The major precondition on
the use of these approaches is, at the minimum, a specification of all
the relevant aspects of the system under development.  Such a
specification enables simple verification techniques such as runtime
checking (via assertions and design by contract~\cite{Meyer1997} using
preconditions, postconditions and invariants) and the automatic
generation of test suites~\cite{Cheon2002,Cheon2004,Cheon2005}.  Via a
complete and sound system specification, supported by an appropriate
formal language (to specify the system model and its properties)
coupled with a (possibly restricted) programming language, one can
statically prove properties about the system using a variety of tools
and
techniques~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

% previous work: why is not satisfying
The underlying idea of this approach is simple: what was once the
purpose of handwritten tests (verification, documentation, and
requirement specification) are now the responsibility of the formal
specifications, and a complete formal specification of the system
under development is needed to apply these methodologies.  The problem
is that this requirement is very difficult to fulfill since the effort
required to write a complete formal specification of the system in
most real-world complex cases is usually greater than writing a suite
of tests~\cite{Glass2002}.

If a complete formal specification is not available, all of the
previous methodologies share a common problem: the parts informally
specified are not verified at all.  These parts of the system are
consequently verified with traditional methods, but the connections
between development artifacts and related activities (code, tests,
design, refactoring) are neither detailed nor enforced.  This is
problematic, as the loops and relations between the different
development artifacts are the inner engine of many AMs, as they impose
an iterative and incremental pace to the development process.

% outsider
Liu takes a different approach wherein the SOFL
methodology~\cite{Liu2004} is merged into agile
processes~\cite{Liu2010}.  The SOFL methodology is based on a three
step specification approach: (1) informal specification, (2)
semi-formal specification, (3) formal specification.  The agile
adaptation consists mainly in introducing shorter loops between
activities and reducing the size of the set of formal specifications.
The formal specifications are used only to help understand ambiguous
statements in the semi-formal specification and are not maintained
while the system evolves.  Testing and inspections are based on the
semi-formal specification.  The formal specification are partial and
are only used for documentation purposes and then discarded---they are
not used at all in the verification of the system.

% what we want to achieve
In contrast, our objective is to formally specify only those parts of
the system that must be verified (in order to cope with constrained
resources and unstable requirements) and to develop and refine a
highly iterative and incremental development process that blends
formal and agile practices.

% how we want to achive it
The main open problem unaddressed in the literature is how to connect
development artifacts of the two worlds of FMs and AMs together.  To
answer this challenge we define activity cycles, similar to what is
found in AMs (especially in XP), to solve this problem in an highly
iterative development process.  We show how tests drive the formal
specification, how the formal specifications drive test and code
development, how handmade and automated tests coexist and support each
other, how the unspecified parts of the system are incrementally
specified.  That is, we show how to blend, and not just tack together,
formal validation and agile verification practices: a \mole of
verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly
  spiced Mexican sauce made chiefly from chili peppers (agile) and
  chocolate (formal), served with meat (working system).}.

@@

\mole practices are applied in the context of a real world case study that matches the problem requirements previously detailed: unstable requirements and development artifacts that need to be formally specified and validated.

@@

% device drivers and formal specification
When considering the development of the device and of its software driver as separate entities that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware-software co-design~\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}, all of which focus on a specification that aims to be as complete as possible.
A complete specification is neither feasible nor convenient in our case. 
The protocol specification must be enhanced incrementally and its complete specification cannot be provided.  
This makes the problem an ideal candidate to test our \mole of practices.

@@

\subsection{\mole Practices}

@@

% all together blends into a mole
When the \mole verification practices are used together the initial objectives are achieved: an incremental and iterative process, where development pace is defined by very short development cycles similar to TDD; the complete freedom to decide what is formally specified; the development cycles and the associated verification practices maintain the corresponding artifacts involved consistent, the correct development cycle is determined by the formality of the artifacts involved, ranging from pure development cycles, shown in Figure \ref{fig:tdd_diagram} and \ref{fig:fdd_diagram}, to blended ones, shown in Figure \ref{fig:fdd_hand_made_tests_diagram} and \ref{fig:tdf_diagram}.

% =====================================================================
\section{\mole Verification Practices Applied}
\label{sec:a_complex_case_study}

\subsection{Data Stream, Protocol Description and Protocol Specification}
\label{subsec:data_stream_and_protocol_description}

@@

% verification
Verification is provided by handmade test suite, JML annotations and generated unit test suite. 
It has been built starting from handmade tests only (TDD development cycles).
Later on JML annotations have been added incrementally, using a mix of \mole practices: to move from handmade tests to formal specifications, TDF have been initially used; than all the \mole practices have been used according to the artifacts involved.

% development cycles
The test suites are used to verify board simulators (high-level), RAC and JML annotations are used to check the formally specified properties at runtime.
Code and tests have been initially implemented using TDD, than using all the \mole verification practices.

@@

\subsection{JML Specification Examples}
\label{subsec:a_jml_specification_example}

@@

\subsection{Test Cases}
\label{subsec:test_cases}

@@

\subsection{Test Cases on the Job}
\label{subsec:test_cases_on_the_job}

@@

\section{Retrospective on \mole Practices Effectiveness}
\label{subsec:test_cases_retrospectives}

% preconditions on projects
We believe that \mole practices are relatively easy to introduce and
use.  Many kinds of projects could benefit from these practices, not
only those that have unstable requirements but also demands for a
certain level of formality.  Even projects with no requirements for
formal specification at all benefit from these practices, since some
simple or obvious system properties are easy to specify and, though
their introduction, there is a corresponding reduction in the test
suite development effort.

% precondition on formal methodology and associated formal suite
The cornerstone of the practices is the formal methodology and its
supporting tools, and the precondition on the tools' utility is not
always easily fulfilled.  Methodologies that require a complete
formalization are nearly impossibile to apply.  In addition, it is not
clear if a tool suite that does not provide a runtime verification,
either via runtime checking or simulation, can be broadly used
succesfully, nor it is clear whether formal methods that are not based
on the contract paradigm are useful in this context.  These are open
questions that will need to be addressed in the future to determine
the broad, concrete applicability of the \mole practices.

% when to use what
There are no strict rules to determine which of the \mole practices to
use when facing a specific problem.  We have found that experience
trumps cookbooks, so we can only give some advice.  We suggesting
using TDD there is no need for any specifications at all, FDD with
generated tests if the specification are not too difficult and system
scenarios is simple enough to be covered by generated tests, FDD with
hand made tests if the scenarios are complex, and TDF when the
specification is complex.  We feel that the experience we have
gathered is not yet enough to distill richer idioms or patterns---such
is future work.

% what is working and what is not
Upon reflect, we find that the end result obtained via combining FDD
with hand made tests is remarkable, as it is as natural to work with
as TDD.  FDD with generated tests is sometimes annoying because of the
time needed to run the full test suite, thus often only part of the
test suite is executed so that the development loops do not take too
long.  TDF is great for complex preconditions, but really difficult
with postconditions, especially if the specification is written for
existing code.  When writing after code, the corresponding test is not
expected to fail, hence the feedback is limited and the contract is
often not as strong as needed.  In this specific case, TDF
effectiveness is limited.

@@

% the problem
In this paper we focus on a specific set of development challenges
with which neither AMs nor FMs are completely confortable.  The
challenges are characterized by unstable requirements combined with
artifacts that must be formally specified and verified all while the
development team is constrained by deadlines that cannot be postponed.
We propose a blend of agile and formal engineering practices, enclosed
by an iterative and incremental development process: the \mole
development process.

% Mole
The verification practices described in this paper recreate the fast
feedback development environment we find in Test Driven Development in
the presence of formal and informal artifacts.  A total of four
practices are presented---two of them are conservative, following
closely the development practices of formal methodologies and agile
methodologies, while the other pair are innovative, blending together
elements from both worlds.

% Mole at work
The practices have been applied succesfully to develop a driver for a
custom embedded sensor board equipped with board simulators and
protocol verifiers.  And, while the protocol and the board
specifications were informal and unstable, a working software product
was needed as soon as possible, with incrementally added
functionality, and ensuring that the software and hardware products
were linked via the last specifications of the protocol and the board.
\mole verification practices enabled us to keep in sync formal
specifications (JML annotations), informal specifications (test
suites), and source code through very rapid, short development cycles.
We developed handmade test suites and generated test suites and JML
annotations, all of which successfully supported the verification of
the software driver and the simulators, as well as the hardware board
when it was delivered.  In fact, the resulting system had no flaws, as
it worked correctly the very first time it was plugged in to the
prototype hardware.

% extending Mole
The practices we describe are general and are applicable to different
development languages and formal methodologies.  One important
constraint is on the formal methodology is that it must cope with
partial specifications.  Another constraint is on the tools available,
which must verify system properties at runtime and partially automate
the verification process of the formalized properties.

TODO POP!

@@

%=====================================================================
% hi Joe

% what about having the chat next week?
% on thursday, maybe?

% here are my comments on the reviewers:

% ---------------------------- REVIEW 1 --------------------------
% OVERALL RATING: 3 (strong accept)

% > This paper successfully bridges the gab between both worlds. I would be i=
% nteresting to read at bit more about the gained experience with the techniq=
% ues based on the presented case study (what was easy what difficult).

% agree. to be included

% > Furthermore I recommend that the authors work out a bit more detailed the=
%  prerequisites to apply a mixture of formal and agile techniques (e.g. by m=
% eans of a list of "thesis"). This would stimulate the discussion!

% agree. to be included. it's already there, but not clear and polished enoug=
% h

% ---------------------------- REVIEW 2 --------------------------
% OVERALL RATING: 2 (accept)

% > The authors use the (somewhat formal) JML technology and introduce their =
% TDF-methodology as a variant of more classic development cycles.

% why "somewhat"?

% > Content-wise it must be said that, given the amount of material in the te=
% xt, the conclusions drawn and lessons learnt are a bit 'thin'. E.g. Does TD=
% F really work? what is your experience?

% partial agree. to be expanded, in the line of the previous comments

% > It seems that the format of the printed area (text height and width) is n=
% ot in accordance with the other conference contributions. When reformatted,=
%  the already long paper will probably get longer. The (too large) wordiness=
%  of the paper could perhaps be reduced somewhat to compensate.

% uh
% I used thir template... I'm "almost" sure
% I'll double check this
% maybe some following format options

% > page 1, abstract: 'not always see eye to eye' seems an inappropriate way =
% of stating 'do not meet face to face'.
% > page 1, abstract end: highly-spices -> highly-spiced
% > page 2, line 2: citation problem
% > page 2, line 12: 'ther'?
% > page 3, line 8 and 10: appliable in -> applicable to?
% > page 3, line 8: yeilding -> yielding
% > page 3, middle: .... the handmade test suites are automatically generated=
%  .... ?
% > page 3, the same section: the last two sentences seem crippled; split the=
% m up!
% > page 6, line -5: Identify -> Identifying
% > page 9, line 1: confidant -> confident
% > page 15, line 6: inoovative -> innovative
% > page 15, line 12: in sink -> in sync
% > page 17, last line: litt ref in disorder

% ok

% > Summing up the editorial part: Please use a spelling checker and perhaps =
% some review by a native speaker
% could help sharpening (and shortening) some sentences.

% well... I've used a spell checker!
% but we have to reread it once the modifications are in place

% ---------------------------- REVIEW 3 --------------------------
% OVERALL RATING: 2 (accept)

% > they do not mention Liu's work on the SOFL foraml engineering method for =
% software development that have been xtensively studies over the last 15 yea=
% rs with the aim to solve the similar problem. In particular, the recent pap=
% er on the application of SOFL to achieve agile process has provided a very =
% different approach from the mentioned approaches
% > Shaoying Liu, ``An Approach to Applying SOFL for Agile Process and Its Ap=
% plication in Developing a Test Support Tool'', Innovations in Systems and S=
% oftware Engineering, Springer London, 22 December, DOI 10.1007/s11334-0114-=
% 3.

% ok
% to be read and analyzed

% ---------------------------- REVIEW 4 --------------------------
% OVERALL RATING: 1 (weak accept)

% > I did find it disturbing that the paper never clearly distinguishes betwe=
% en structural and functional testing . This is an
% important distinction as it potentially evidences different types of proble=
% ms.

% I'll rewrite some parts to make the distinction clear

% > Moreover, there is a distinct difference between automatically generated =
% tests and manually produced tests as to how well they apply to one or the o=
% ther.

% the distinction is clear in the paper... I don't understand this

% > The overall description of how AM and FM are merged is not that persuasiv=
% e, as a result. In particular, Figure 2 does not
% convey much nor profound value in its current incarnation and description.

% mmhh
% not much that can be done here... I think
% how can we be more "persuasive"?

% > It would help if the authors would more specifically address what type of=
%  tests the JML specification helps generate. In the
% case study, it is a structural coverage type test.

% mmhh
% I've explained this
% it would take me pages to go into specific details... and I've cited
% works where this is explained

% > All that said, it is a bit unclear as to where the presented work is actu=
% ally pushing the boundaries of the current state of
% the field, other than in actually presenting the approach as an FM+AM hybri=
% d.

% we are actually talking about FM+AM hybrids... that IS the filed,
% so... what's this question about?

% > As a suggestion, then, it could be good if the verbosity and obtusity of =
% the article is reduced

% ok...

% > and a more formal and incisive treatise is presented.

% no idea how to do this
% nor I'm sure I really want to do this

% --=20
% Vieri del Bianco
